---
title: "Homework 1 (2023): Data Wrangling and Visualization"
subtitle: "Due date : 2023-02-24 23:55 (this is a hard deadline)"
author: "Lagarde et Michard"
date: "2023-02-24"
format:
    html:
        code-fold: true
editor: 
  markdown: 
    wrap: 72
---

# Homework 1

Due date : **2023-02-24 @23h55** (this is a **hard deadline**)

## Fill this with your names

-   AVOSSE, Yaovi Ametepé, Data
-   EL BAZ, Reda, MFA

## Carefully follow instructions

**If you don't: no evaluation!**

Write in English or French

The deliverable is a file

-   `xxx_yyy.Rmd` file (Rmarkdown) or
-   `xxx_yyy.qmd` file (if you are using `quarto`)

where `xxx` and `yyy` are your names, for example
`lagarde_michard.ipynb`.

The deliverable is not meant to contain cell outputs.\
The data files used to execute cells are meant to sit in the same
directory as the deliverable. Use relative filepaths or urls to denote
the data files.

I **will** execute the code in your notebook: make sure that running all
the cells works well.

## Grading <i class="fa graduation-cap"></i>

Here is the way we'll assess your work

| Criterion                  | Points | Details                                                                             |
|:-----------------------------|:-----------------------:|:----------------|
| Spelling and syntax        |   3    | English/French                                                                      |
| Plots correction           |   3    | Clarity / answers the question                                                      |
| Plot style and cleanliness |   3    | Titles, legends, labels, breaks ...                                                 |
| Table wrangling            |   4    | ETL, SQL like manipulations                                                         |
| Computing Statistics       |   5    | SQL `goup by` and aggregation                                                       |
| DRY compliance             |   2    | DRY principle at [Wikipedia](https://en.wikipedia.org/wiki/Don%27t_repeat_yourself) |

If we see a single (or more) `for` loop in your code: **-5 points**.
Everything can be done using high-level `pan
 das` methods

# Preliminaries

## Notebooks, Rmarkdown, Quarto: Modus operandi

-   [Quarto](https://quarto.org)
-   [Rmarkdown](https://rmarkdown.rstudio.com)
-   [Jupyter for R](https://jupyter.org)

## Packages

-   Base `R` can do a lot. But the full power of `R` comes from a fast
    growing collection of \`packages\`\`.

-   Packages are first installed, and if needed, imported during a
    session.

-   Once a package has been installed on your drive, if you want all
    objects exported by the package to be available in your session, you
    should import the package, using `require(...)` or `library()`.

-   If you just want to pick some subjects from the package, you can use
    qualified names like `pkgname::object_name` to access the object
    (function, dataset, class...)

```{r}
#pacman::p_load("tidyverse")
if (! "pacman" %in% installed.packages()[,1])
  install.packages("pacman")

p_load <- pacman::p_load

p_load("tidyverse")

pacman::p_load(forcats,
               Rcurl,
               zip,
               httr,
               dplyr,
               readr,
               utils,
               utf8,
               vroom,
               readxl)
```

# Getting the data

## French data

The French data are built and made available by
[INSEE](https://www.insee.fr/fr/accueil) (French Governement Statistics
Institute)

Prénoms: -
<https://www.insee.fr/fr/statistiques/fichier/2540004/nat2021_csv.zip>

This dataset has been growing for a while. It has been considered by
social scientists for decades. Given names are meant to give insights
into a variety of phenomena, including religious observance.

-   A glimpse at the body of work can be found in [*L'archipel français*
    by Jérome Fourquet, Le Seuil,
    2019](https://www.seuil.com/ouvrage/l-archipel-francais-jerome-fourquet/9782021406023)

-   Read the [File
    documentation](https://www.insee.fr/fr/statistiques/2540004?sommaire=4767262#documentation)

## US data

US data may be gathered from

[Baby Names USA from 1910 to 2021
(SSA)](https://www.kaggle.com/datasets/donkea/ssa-names-1910-2021?resource=download)

See <https://www.ssa.gov/oact/babynames/background.html>

## British data

English and Welsh data can be gathered from

[https://www.ons.gov.uk/](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/datasets/babynamesinenglandandwalesfrom1996?utm_source=pocket_saves)

## Download the French data

**QUESTION:** Download the data into a file which relative path is
`'./nat2021_csv.zip'`

**Hints:**

-   Have a look at package
    [`httr`](https://cran.r-project.org/web/packages/httr/index.html).
-   Use magic commands to navigate across the file hierarchy and create
    subdirectories when needed

```{r}
# for French data 

params <-  list(
    url = 'https://www.insee.fr/fr/statistiques/fichier/2540004/nat2021_csv.zip',
    dirpath = './',
    timecourse = '',
    datafile = 'nat2021.hdf',
    fpath = 'nat2021_csv.zip'
)
```

```{r}
# modify location  make sure you are in the right directory
 wd <-getwd() # get working directory
  # set working directory
# getwd() 
# list.dirs(). # list files in working directory
```

```{r}
# your code here
url <- "https://www.insee.fr/fr/statistiques/fichier/2540004/nat2021_csv.zip"
data_file <- tempfile(file = ".zip")
download.file(url, data_file, mode = "wb")

```

## Download US and British data

```{r}
# your code here
# Download US data
url_US <- "https://www.ssa.gov/oact/babynames/names.zip"
data_file_US <- tempfile(file = ".zip")

download.file(url_US, data_file_US, mode = "wb")

unzip(data_file_US,
      overwrite = TRUE,
      exdir = "./babies-us")
list_babynames_US <- list.files(path = "./babies-us",
                                pattern = ".txt",
                                full.names= TRUE)
```
```{r}
url_brit <- "https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/datasets/babynamesinenglandandwalesfrom1996/1996tocurrent/babynames1996to2021.xlsx"
data_file_brit <- tempfile( fileext=".xslx")
download.file(url_brit, data_file_brit, mode = "wb")
```

## Load the French data in memory

**QUESTION:** Load the data in a `pandas` `DataFrame` called `data`

**Hints:**

-   You should obtain a `dataframe`/`tibble` with 4 columns.
-   Mind the conventions used to build the `csv` file.
-   Package `pandas` provides the convenient tools.
-   The dataset, though not too large, is already demanding.
-   Don't hesitate to test your methods on a sample of rows. Function
    `dplyr::slice_sample()` can be helpful.

```{r}
# your code here
# Load french data
baby_names_french <- read_csv2(data_file)
baby_names_french

```

## Load US and British data in memory

```{r}
# your code here
# Load US data
baby_names_US <- data.frame()
for (i in 1:142) {
      dfi <- vroom(list_babynames_US[i],
                   delim=",",
                   col_names = c("preusuel","gender","nombre"),
                   show_col_types = FALSE)
      
      dfi$annais <- rep(1879+i,nrow(dfi))
      baby_names_US <- rbind(baby_names_US,dfi)
}
baby_names_US
```

```{r}
#Load British Data
baby_names_british_M <- read_xlsx(data_file_brit, sheet = 4, col_names = TRUE, skip= 7)
baby_names_british_F <- read_xlsx(data_file_brit, sheet = 5, col_names = TRUE, skip= 7)

```

## Explore the data
  
**QUESTION:** Get a feeling of the data.

-   This dataset is supposed to report all given names used for either
    sex during a year in France since 1900

-   The file is made of `> 600 000` lines and 4 columns.

```{=html}
<!-- -->
```
    |-- preusuel : object
    |-- nombre: int64
    |-- sexe: int64
    |-- annais: object

Each row indicates for a given `preusuel` (prénom usuel, given name),
`sexe` (sex), and `annais` (année naissance, birthyear) the `nombre`
(number) of babies of the given sex who were given that name during the
given year.

| sexe | preusuel | annais | nombre |
|:-----|:---------|-------:|-------:|
| 2    | SYLVETTE |   1953 |    577 |
| 1    | BOUBOU   |   1979 |      4 |
| 1    | NILS     |   1959 |      3 |
| 2    | NICOLE   |   2003 |     36 |
| 1    | JOSÉLITO |   2013 |      4 |

**QUESTION:** Compare memory usage and disk space used by data

```{r}
# your code here
# memory usage
message("memory usage:")
object.size(baby_names_french) 
# for memory space
message("memory space:")
file.size("./nat2021.csv")
```
```{r}
# for US data
# memory usage
message("memory usage: ")
object.size(baby_names_US) 
# for memory space
total_size_file_US <- 0
for (i in 1:142) {
  total_size_file_US <- total_size_file_US + file.size(list_babynames_US[i])
}
message("memory space:")
total_size_file_US
```
**QUESTION:** For each column compute the number of distinct values.

```{r}
# your code here
# for french data
baby_names_french %>%
  summarise(distinct_sexe = n_distinct(sexe),
            distinct_preusuel = n_distinct(preusuel),
            distinct_annais = n_distinct(annais),
            disctinct_nombre = n_distinct(nombre)
            )
```
```{r}
# for US data
# your code here
baby_names_US %>%
  summarise(distinct_gender = n_distinct(gender),
            distinct_firstname = n_distinct(firstname),
            distinct_year = n_distinct(year),
            disctinct_count = n_distinct(count)
            )
```

# Transformations

## Improving the data types

**QUESTION:** Make `sexe` a category with two levels `Female` and
`Male`. Call the new column `gender`. Do you see any reason why this
factor should be ordered?

Use package `forcats`

```{r}
# your code here
baby_names_fr <- baby_names_french %>%
    mutate(gender=ifelse(sexe==1, "Male","Female"))%>%
    mutate(gender= as_factor(gender))
baby_names_french
```

**QUESTION:** Compare memory usage of columns `sexe` and `gender`

```{r}
# your code here
# memory usage of sexe:
object.size(baby_names_fr$sexe)

# memory usage of gender:
object.size(baby_names_fr$gender)

# We remark that memory usage of sexe is greater than memory usage of gender
```
```{r}
# your code here
#US Data
# memory usage of sexe:
object.size(baby_names_US$sexe)

# memory usage of gender:
object.size(baby_names_US$gender)

# We remark that memory usage of sexe is greater than memory usage of gender
```


**QUESTION:** Would it be more memory-efficient to recode `sexe` using
modalities `F` and `M` instead of `Male` and `Female` ?

*Insert your answer here*

> Using modalities 'F' and 'M' instead of 'Male' and 'Female' doesn't affect the momory-efficient

## Dealing with missing values

**QUESTION:** Variable `annais` class is `object`. Make `annais` of type
`numeric`. Note that missing years are encoded as "XXXX", find a way to
deal with that.

```{r}

# your code here
baby_names_french <- baby_names_french %>%
    drop_na() %>%
    mutate(annais=as.numeric(annais))%>%
    filter(annais !='XXXX')

```

## Rename and remove columns

**QUESTION:** Remove useless columns (now that you've created new ones,
and rename them). You should end up with a dataframe with columns called
`"gender"`, `"year"`, `"count"`, `"firstname`" with the following
dtypes:

```{verbatim}
gender        factor
firstname     character
count         integer
year          integer
```

```{r}
# your code here
baby_names_french<-  baby_names_french %>%
                      mutate(year=as.integer(annais),
                      count=as.integer(nombre),
                      firstname= preusuel) %>%
                      select(-sexe, -annais,-nombre,-preusuel)
baby_names_french
```

**Question:** Do the same thing for British and US data. You should
eventually obtain dataframes with the same schema.
```{r}
# your code here
baby_names_US<-  baby_names_US %>%
                      mutate(year=as.integer(annais),
                      count=as.integer(nombre),
                      firstname= preusuel) %>%
                      select(-annais,-nombre,-preusuel)
baby_names_US
```

**QUESTION:** How many missing values (NA) have been introduced? How
many births are concerned?

```{r}
# your code here
baby_names_US %>% 
  summarise(across(everything(), ~  sum(is.na(.)))) %>% 
  pivot_longer(cols = everything()) %>% 
  filter(value > 0) %>% 
  arrange(desc(value))
```
```{r}
baby_names_french %>% 
  summarise(across(everything(), ~  sum(is.na(.)))) %>% 
  pivot_longer(cols = everything()) %>% 
  filter(value > 0) %>% 
  arrange(desc(value))
```

**QUESTION:** Read the documentation and describe the origin of rows
containing the missing values.

```{r}

```

## Checkpointing: save your transformed dataframes

**QUESTION:** Save the transformed dataframe (retyped and renamed) to
`./nat2021_csv.zip`. Try several compression methods.

```{r}
# your code here
baby_names_french %>% 
  write.csv(file = "./babies-fr.csv")
zip(zipfile="./babies-fr.zip", files="./babies-fr.csv")

```

```{r}
baby_names_US %>% 
  write.csv(file = "./babies-us.csv")
zip(zipfile="./babies-us.zip", files="./babies-us.csv")
```

**QUESTION:** Save the transformed dataframes (retyped and renamed) to
`./nat2021.hdf` using `.hdf` format

```{r}
# your code here
```

At that point your working directory should look like:

    |── homework01.rmd.    # if you use `rmarkdown`
    |── homework01.qmd     # if you use `quarto`
    ├── homework01.ipynb   # if you use `jupyter` `notebook`
    ├── babies-fr.zip
    ├── babies-us.zip
    ├── babies-ew.zip
    ├── births-fr.csv

**QUESTION:** Reload the data using `read_hdf(...)` so that the
resulting dataframes are properly typed with meaningful and homogeneous
column names.

**Hint:** use `tryCatch(...)` to handle exceptions

```{r}
# your code here
```

## Some data "analytics" and visualization

**QUESTION**: For each year, compute the total number of Female and Male
births and the proportion of Female births among total births

**Hints:**

```{r}
# your code here
# for female
Female_birth_per_year<- baby_names_french %>% 
                             group_by(year)%>% 
                             filter(gender=="F")%>% 
                             summarise(total_female_birth_fr = sum(count))


# for male
Male_birth_per_year<- baby_names_french %>% 
                             group_by(year)%>% 
                             filter(gender=="M")%>% 
                             summarise(total_male_birth_fr = sum(count))

Births_per_year_fr <- merge(Female_birth_per_year, Male_birth_per_year)
Births_per_year_fr
```

```{r}
# your code here
# for female
Female_birth_per_year_US<- baby_names_US %>% 
                             group_by(year)%>% 
                             filter(gender=="F")%>% 
                             summarise(total_female_birth_US = sum(count))

# for male
Male_birth_per_year_US<- baby_names_US %>% 
                             group_by(year)%>% 
                             filter(gender=="M")%>% 
                             summarise(total_male_birth_US= sum(count))

Births_per_year_US <- merge(Female_birth_per_year_US, Male_birth_per_year_US)
Births_per_year_US
```


**QUESTION:** Plot the proportion of female births as a function of year
and French, US, en British babynames data. Compare with what you get
from `births-fr.hdf`.

Don't forget: title, axes labels, ticks, scales, etc.

Because of what we did before, the `plot` method of a `DataFrame` with
be rendered using `plotly`, so you can use this. But you can use also
`seaborn` or any other available plotting library that you want.

**Hint:** Mind the missing values in the `year` column

```{r}
# your code here
Births_per_year <- merge(Births_per_year_fr, Births_per_year_US)
# female proportion per year

F_births_proportion_fr <- Births_per_year %>%
  reframe(year, proportion_female_birth_fr = total_female_birth_fr / (total_female_birth_fr + total_male_birth_fr))

F_births_proportion_US <- Births_per_year %>%
  reframe(year, proportion_female_birth_US = total_female_birth_US / (total_female_birth_US + total_male_birth_US))

F_births_proportion <- merge(F_births_proportion_fr, F_births_proportion_US)
F_births_proportion
```
```{r}
F_births_proportion %>%
  ggplot() +
  aes(x= year)+
  geom_line(aes( y= proportion_female_birth_fr), color= "blue")+
  geom_line(aes( y= proportion_female_birth_US), color= "red")+
  xlab("Years")+
  ylab("Percentage")

```

**QUESTION:** Make any sensible comment about these plots.

*Insert your answer here*

> There is a constant decrease in the proportion of girls being born in both countries with a large drop happening around 1910, it has now stabilized around .48 in both countries after 1950

**QUESTION:** Explore the fluctuations of sex ratio around its mean
value since 1945 in the US, in France and in the Great Britain.

Plot deviations of sex ratio around its mean since 1945 as a function of
time.

```{r}
# your code here
# for french data
# fluctuation ratio of sex ratio around its mean value since 1945
french_fluctuation_ratio <- subset(baby_names_french, baby_names_french$year >= 1945)

french_fluctuation_ratio <- french_fluctuation_ratio %>%
  group_by(year)%>%
  summarise(ratio_gender = mean(french_fluctuation_ratio$count))
french_fluctuation_ratio

# Je n'ai pas encore fini ceci...

```

**QUESTION:** Assume that baby gender is chosen at random according to a
Bernoulli distribution with success probability $.48$, that baby genders
are i.i.d. Perform simulations for sex ratios for French and US data
since 1945.

Plot the results, compare with your plots above.

# The rise, decline and fall of firstnames

**Question:** For each year, country, gender and firstname, compute the
popularity rank of the firstname among the names given to babies with
that gender, in that country, in that year. The most popular name should
be given rank $1$.

**QUESTION:** For each firstname and sex (some names may be given to
girls and boys), compute the total number of times this firstname has
been given during `1900-2019`. Print the top 20 firstnames given and
style your result dataframe using `DT::datatable()` for instance.

```{r}
# your code here
```

## Rare firstnames

**QUESTION:** In the French data, for each sex, plot the proportion of
births given `_PRENOMS_RARES` as a function of the year.

```{r}
# your code here
```

# A study of the "Marie" firstname

**QUESTION:** Plot the proportion of female births given name `'MARIE'`
or `'MARIE-...'` (compounded names) as a function of the year. Proceed
in such a way that the reader can see the share of compounded names. We
are expecting an *area plot*.

**Hints:**

-   Have a look at the `.str` accessor (to apply a string method over a
    whole column containing string)

-   Have a look at [r-graph-gallery: stacked
    area](https://www.r-graph-gallery.com/stacked-area-graph.html) and
    at [ggplot
    documentation](https://ggplot2.tidyverse.org/reference/geom_ribbon.html).
    Pay attention on the way you stack the area corresponding to names
    matching pattern 'MARIE-.\*' over or under the are corresponding to
    babies named 'MARIE'

-   See Graphique 3, page 48, de *L'archipel français* de J. Fourquet.
    Le Seuil. Essais. Vol. 898.

-   Add annotation, 1st World War, Front Populaire, 2nd World War, 1968

```{r}
# your code here
```

# Top 10 firstnames of year 2000

**QUESTION:** For each sex, select the ten most popular names in year
2000, and plot the proportion of newborns given that name over time.
Take into account that some names might have zero occurrence during
certain years.

**Hint:** Leave aside the rows with '\_PRENOMS_RARES'.

```{r}
# your code here
```

# Picturing concentration of babynames distributions

Every year, the name counts define a discrete probability distribution
over the set of names (the universe).

This distribution, just as an income or wealth distribution, is
(usually) far from being uniform. We want to assess how uneven it is.

We use the tools developed in econometrics.

Without loss of generality, we assume that we handle a distribution over
positive integers $1, \ldots, n$ where $n$ is the number of distinct
names given during a year.

We assume that frequencies $p_1, p_2, \ldots, p_n$ are given in
ascending order, ties are broken arbitrarily.

The `Lorenz function`
([Lorenz](https://en.wikipedia.org/wiki/Lorenz_curve) not `Lorentz`)
maps $[0, 1] \to [0, 1]$.

$$L(x) = \sum_{i=1}^{\lfloor nx \rfloor} p_i .$$

Note that this is a piecewise constant function.

**Question:** Compute and plot the Lorenz fucntion for a given `sex`,
`year` and `country`

```{r}
Lorenz <- function(x,sex,year2, country){
S <- country %>%
  filter(firstname != "_PRENOMS_RARES") %>%
  filter(year==year2 & gender==sex) %>%
  summarise(Tot=sum(count))
n <- country %>%
  filter(firstname != "_PRENOMS_RARES") %>%
  filter(year==year2 & gender==sex) %>%
  summarise(n_distinct(firstname))

TOPX <- country %>%
  filter(firstname != "_PRENOMS_RARES") %>%
  filter(year==year2 & gender==sex) %>%
  group_by(firstname) %>%
  mutate(freq=count/S) %>%
  arrange(freq) %>%
  head(floor(n[[1,1]]*x))

L <- sum(TOPX$freq)
return(L)
}

```


```{r}
x <-seq(0,1,0.05)
List_Lorenz <- lapply(x,FUN=Lorenz,"F","1920",baby_names_french)
plot(x,List_Lorenz, type="l")
```


**Question:** Design an animated plot that shows the evolution of the
Lorenz curve of babynames distribution through the years for a given sex
and country.

```{r}
```

The Lorenz curve summarizes how far a discrete probability distribution
is from the uniform distribution. This is a very rich summary and it is
difficult to communicate this message to a wide audience. People tend to
favor numerical indices (they don't really understand, but they get used
to it): Gini, Atkinson, Theil, ...

The [Gini index](https://en.wikipedia.org/wiki/Gini_coefficient) is
twice the surface of the area comprised between curves $y=x$ and
$y=L(x)$.

$$G = 2 \times \int_0^1 (x -L(x)) \mathrm{d}x$$

The next formula allows us to compute it efficiently.

$$G={\frac {2\sum _{i=1}^{n}i p_{i}}{n\sum _{i=1}^{n}p_{i}}}-{\frac {n+1}{n}}.$$

**Question:** Compute and plot Gini index of names distribution over
time for sex and countries

```{r}
# your code here
```

# Picturing surges of popularity

In the sequel, the *popularity* of a gendered name in a population
during a given year is the proportion of babies of that gender born
during that year in that country, that are given this name.

**Question:** Prepare a data frame that contains for each hype name the
20 years before and 30 years after the maximum popularity is achieved,
and, for each such year, the rank and popularity of the hype name. Do
this for US and French data.

```{r}
# your code here
```

**Question:** Plot offseted popularity (share of given names within
year, country, gender) curves of hype names. Facet by sex and country.

```{r}
# your code here
```

**Question:** Rescale popularity curves so that all of them have maximum
$1$.

```{r}
```

# Getting help

-   [tidyverse](https://tidyverse.org)

-   [plotly](https://plotly.com/R/) for animated plots

-   [stackoverflow](https://stackoverflow.com)

-   <https://github.com/hadley/babynames>

-   Don't Repeat Yourself (DRY) principle at
    [Wikipedia](https://en.wikipedia.org/wiki/Don%27t_repeat_yourself)
